Inferring stellar rotation periods with Gaussian processes

### 2016-11-17 documents/ms.tex
adding all code from /GProtation/code/tests

\begin{abstract}
The light curves of spotted, rotating stars are often non-sinusoidal and
quasi-periodic---spots are not static on the stellar surface and have finite
lifetimes causing stellar flux variations to slowly shift in phase.
A strictly periodic sinusoid is therefore not a representative generative
model for the light curve of star with a rotational signal.
Ideally, a physical model of the stellar surface would be conditioned on the
data, however the parameters of such models can be highly degenerate
\citep[\eg][]{Russell1906, Jeffers2009}.
Instead, we use an appropriate {\it effective} model: a Gaussian Process (GP)
with a quasi-periodic covariance kernel function.
By modelling the covariance matrix of the light curve with a GP, a highly
flexible semi-parametric function, we avoid the necessity to choose a `best
fitting' functional form, whilst sampling directly from the posterior
probability distribution function of the periodic parameter and marginalising
over the other kernel hyperparameters.
We used \nlightcurves\ simulated light curves with a range of rotation periods
to test the GP model.
We attempted to recover the rotation periods using three methods: our GP
method, a sine-fitting periodogram method and an autocorrelation function
method.
The posterior probability distribution of the rotation period parameter was
sampled using the affine invariant ensemble MCMC sampler {\tt emcee}
\citep{Foreman-Mackey2013}, and the GP operations were performed using the
{\tt george} python package \citep{George}.
Rotation periods inferred via this method are more precise and accurate than
the periodogram and ACF methods.
Furthermore, the improvement is expected to be even more dramatic when applied
to real, noisy {\it Kepler} light curves, since the GP method is well suited
to modelling rotation signals and correlated noise simultaneously.

@@

The light curves of spotted, rotating stars are often non-sinusoidal and
Quasi-Periodic (QP).
These stars vary in brightness due to active regions on their surfaces which
rotate in and out of view.
The non-sinusoidal quality is caused by the complicated surface spot patterns
and the quasi-periodicity is produced both by the finite lifetimes of these
active regions and the presence of differential rotation on the stellar
surface.
A strictly periodic sinusoid is therefore not a good model for stellar light
curves.
In an ideal world, a physical model of the stellar surface would be
conditioned on the data.
A physically realistic, generative model would perfectly capture the complexity
of shapes within stellar light curves as well as the quasi-periodic nature,
allowing for extremely precise probabilistic period recovery.
However, such physical models require many free parameters in order to
accurately represent a stellar surface and some of these parameters are
extremely degenerate \citep[\eg][]{Russell1906, Jeffers2009}.
In addition to global stellar parameters such as inclination and rotation
period, each spot or active region should have (at minimum) a longitude,
latitude, size, temperature and lifetime.
Considering that many stars have on the order of hundreds of spots, the number
of free parameters quickly becomes unwieldy, especially if the posterior PDFs
of these parameters are explored with MCMC.
Simplified spot models, such as the one described in \citet{Lanza2014} where
only two spots are modelled, have produced successful results, however these
simplified models sacrifice some precision due to lack of model flexibility.
Instead of using a physical model for stellar light curves, we choose to use
an {\it effective} model: one which captures the behaviour but is not
physically motivated, although the parameters of this model may be {\it
interpreted} as physical ones.
An ideal effective model for the light curve of a spotted, rotating star is
one with a small number of non-degenerate parameters that is flexible enough
to perfectly capture non-sinusoidal, QP behaviour.
These requirements are perfectly fulfilled by a Gaussian process (GP) model.

@@

The motivation for developing a GP-based method for rotation period inference
is, firstly to measure more accurate and precise rotation periods using a
better-suited generative model than a sinusoid for the reasons explained
above.
Secondly, in order to infer {\it probabilistic} periods, i.e. to estimate the
posterior PDF of the rotation period and thereby measure a realistic and
representative uncertainty.
And thirdly, to allow for an additional correlated noise model to be included
during regression, the parameters of which could be marginalised over.
% And fourthly to provide some way of determining whether a periodic model is
% supported by the data over a purely stochastic one.

GPs are commonly used in the machine learning community and increasingly
in other scientific fields, for example biology and geophysics (where GP
regression is called `kriging').
More recently, GPs have been used in the astronomy literature \citep[see
\eg][]{Gibson2012, Haywood2014, Haywood2015, Evans2015, Rajpaul2015,
Rajpaul2016, Aigrain2016}.
% FIXME: CITATIONS Gibson, Aigrain, Evans, Rajpaul, Waldman(?), Cosmology,
% geology, biology, Rasmussen & Williams, etc.
They are useful in regression problems involving any stochastic process,
specifically when the probability distribution for the process is a
multi-variate Gaussian.
If the probability of obtaining a data-set is a Gaussian in $N$ dimensions,
where $N$ is the number of data points, that data-set can be described as, and
with, a Gaussian process.
An in-depth introduction to Gaussian processes is provided by
\citet{Rasmussen2005}.

GP models parameterise the covariance between data points and a kernel
function provides the form of covariance matrix parameterisation.
For example, take the time-series in figure \ref{fig:GP_example}.
% This is a \kepler\ light curve of Earth-like planet host Kepler-452: a G-type
% star that rotates once every $\sim$ 15 days.
This is the \kepler\ light curve of KID 1430163, a relatively active star that
rotates once every $\sim$ 4 days.
The stochastic variability in this time-series is typical of \kepler\ FGK
stars.
Clearly, data points in this light curve are correlated.
Points that are close together in time are tightly correlated and those that
are far apart are loosely correlated.
It is the way in which the covariance varies with the separation between data
points that is modelled when using GP regression.
It is not the data but the {\it covariance structure} that is modelled.
This fact is what gives GPs their flexibility---they can model any time
series with a similar covariance structure.
In addition, a very simple function can usually capture the covariance
structure of a light curve, whereas a much more complicated one may be
required to model the time-series itself.
% The light curve of \Kepler-452 has been modelled with a GP in figure
% \ref{fig:GP_example}, shown in orange.
The light curve of this star has been modelled with two different GP kernel
functions in figure \ref{fig:GP_example}, shown in blue and pink.
Both provide adequate fits to the data, however only the periodic kernel
function, `QP' is a useful model because it has a periodic parameter.
I return to this point shortly.

A range of kernel functions could be used to describe stellar variability.
For example, the simplest and most commonly used kernel function, the `Squared
Exponential' (SE) produces an adequate fit to the KID 1430163 light curve.
The SE kernel function is defined as,
\begin{equation}
k_{i,j} = A \exp \left(-\frac{(x_i - x_j)^2}{2l^2} \right).
\end{equation}
\label{eq:SE}
Here $A$ is the amplitude of covariance, $l$ is the length scale of
exponential decay and $x_i-x_j$ is the separation between data points.
The SE kernel function has the advantage of being very simple---it has just
two parameters, a covariance amplitude and length scale: $A$ and $l$.
If $l$ is large two data points far apart in $x$ will be tightly correlated,
and if small they will be loosely correlated.
Another property of the SE kernel function is that it produces functions that
are infinitely differentiable.
It is therefore possible to model a data set and its derivatives
simultaneously.
% \citep[see \eg][]{Rajpaul2015}
The SE kernel function may be an adequate model of the covariance in stellar
light curves, but it is not a {\it useful} one for the problem of rotation
period inference because it does not have a parameter that describes a period.
In order to infer rotation periods it is necessary to use a periodic kernel
function.
For this reason, we use the `Quasi-Periodic' kernel function.
The QP kernel function is defined as
\begin{equation}
k_{i,j} = A \exp \left(-\frac{(x_i - x_j)^2}{2l^2} -
\frac{\sin^2(\frac{2\pi}{P})}{\Gamma^2} \right).
\end{equation}
\label{eq:QP}
It is the product of the SE kernel function, which describes the overall
covariance decay, and an exponentiated, squared, sinusoidal kernel function
that describes the periodic covariance structure.
$P$ can be interpreted as the rotation period of the star and $\Gamma$
controls the number of zero crossings per period.
This kernel function allows two data points that are separated in time by one
rotation period to be tightly correlated, while points separated by half a
period can be weakly correlated.
Both the QP and SE kernel functions were used to produce the blue and pink
models shown in figure ~\ref{fig:GP_example}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/001430163.pdf}
\caption[A light curve with a GP model.]
{Light curve of KID 1430163, an active star with a rotation period of $\sim$ 4
days.
The blue solid line shows a fit to the data using
a Gaussian process model with a QP covariance kernel function and the pink
broken line shows a fit with a SE kernel function.
Both kernel functions provide an adequate fit to the data, however we use the
QP function because it has a useful parameter that corresponds to the star's
rotation period.}
\label{fig:GP_example}
\end{center}
\end{figure}

>>>>

\begin{abstract}
The light curves of spotted, rotating stars are often non-sinusoidal and
quasi-periodic---spots are not static on the stellar surface and have finite
lifetimes causing stellar flux variations to slowly shift in phase.
A strictly periodic sinusoid is therefore not a representative model for the
light curve of star with a rotational signal.
Ideally, a physical model of the stellar surface would be conditioned on the
data, however the parameters of such models can be highly degenerate
\citep[\eg][]{Russell1906, Jeffers2009}.
Instead, we use an appropriate {\it effective} model: a Gaussian Process (GP)
with a quasi-periodic covariance kernel function.
By modelling the light curve with a GP, a highly flexible semi-parametric
function, we avoid the necessity of choosing a `best fitting' functional form,
whilst sampling directly from the posterior probability distribution function
of the periodic parameter and marginalising over the other kernel
hyperparameters.
We used \nlightcurves\ simulated light curves with a range of rotation periods
to test the GP model.
We attempted to recover the rotation periods using three methods: our GP
method, a sine-fitting periodogram method and an autocorrelation function
method.
The posterior probability distribution of the rotation period parameter was
sampled using the affine invariant ensemble Markov Chain Monte Carlo (MCMC)
sampler {\tt emcee} \citep{Foreman-Mackey2013}, and the GP operations were
performed using the {\tt george} python package \citep{George}.
Rotation periods inferred via this method are more precise and accurate than
the periodogram and ACF methods.
Furthermore, the improvement is expected to be even more dramatic when applied
to real, noisy {\it Kepler} light curves, since the GP method is well suited
to modelling rotation signals and correlated noise simultaneously.

@@

The light curves of spotted, rotating stars are often non-sinusoidal and
Quasi-Periodic (QP).
These stars vary in brightness due to active regions on their surfaces which
rotate in and out of view.
The non-sinusoidal quality is caused by the complicated surface spot patterns
and the quasi-periodicity is produced both by the finite lifetimes of these
active regions and the presence of differential rotation on the stellar
surface.
A strictly periodic sinusoid is therefore not a good model for stellar light
curves.
In an ideal world, a physical model of the stellar surface would be
conditioned on the data.
A physically realistic model would perfectly capture the complexity of shapes
within stellar light curves as well as the quasi-periodic nature, allowing for
extremely precise probabilistic period recovery.
However, such physical models require many free parameters in order to
accurately represent a stellar surface and some of these parameters are
extremely degenerate \citep[\eg][]{Russell1906, Jeffers2009}.
In addition to global stellar parameters such as inclination and rotation
period, each spot or active region should have (at minimum) a longitude,
latitude, size, temperature and lifetime.
Considering that many stars have on the order of hundreds of spots, the number
of free parameters quickly becomes unwieldy, especially if the posterior
Probability Density Functions (PDFs) of these parameters are explored with
MCMC.
Simplified spot models, such as the one described in \citet{Lanza2014} where
only two spots are modelled, have produced successful results, however these
simplified models sacrifice some precision due to lack of model flexibility.
Instead of using a physical model for stellar light curves, we choose to use
an {\it effective} model: one which captures the behaviour but is not
physically motivated, although the parameters of this model may be {\it
interpreted} as physical ones.
An ideal effective model for the light curve of a spotted, rotating star is
one with a small number of non-degenerate parameters that is flexible enough
to perfectly capture non-sinusoidal, QP behaviour.
These requirements are fulfilled by a Gaussian process (GP) model.

@@

The motivation for developing a GP-based method for rotation period inference
is, firstly to measure more accurate and precise rotation periods using a
better-suited generative model than a sinusoid for the reasons explained
above.
Secondly, in order to infer {\it probabilistic} periods, i.e. to estimate the
posterior PDF of the rotation period and thereby measure a realistic and
representative uncertainty.
Thirdly, to allow for an additional correlated noise model to be included
during regression, the parameters of which could be marginalised over.
% And fourthly to provide some way of determining whether a periodic model is
% supported by the data over a purely stochastic one.

GPs are commonly used in the machine learning community and increasingly
in other scientific fields, for example biology and geophysics.
More recently, GPs have been used in the astronomy literature \citep[see
\eg][]{Gibson2012, Haywood2014, Haywood2015, Evans2015, Rajpaul2015,
Rajpaul2016, Aigrain2016}.
% FIXME: CITATIONS Gibson, Aigrain, Evans, Rajpaul, Waldman(?), Cosmology,
% geology, biology, Rasmussen & Williams, etc.
They are useful in regression problems involving any stochastic process,
specifically when the probability distribution for the process is a
multi-variate Gaussian.
If the probability of obtaining a data-set is a Gaussian in $N$ dimensions,
where $N$ is the number of data points, that data-set can be described as, and
with, a Gaussian process.
An in-depth introduction to Gaussian processes is provided by
\citet{Rasmussen2005}.

GP models parameterise the covariance between data points and a kernel
function provides the form of covariance matrix parameterisation.
For example, take the time-series in figure \ref{fig:GP_example}.
% This is a \kepler\ light curve of Earth-like planet host Kepler-452: a G-type
% star that rotates once every $\sim$ 15 days.
This is the \kepler\ light curve of KIC 1430163, a relatively active star that
rotates once every $\sim$ 4 days.
The stochastic variability in this time-series is typical of \kepler\ FGK
stars.
Clearly, data points in this light curve are correlated.
Points that are close together in time are tightly correlated and those that
are widely separated in time are loosely correlated.
The way in which the covariance varies with the separation between data points
is modelled when using GP regression; it is not the data but the {\it
covariance structure} that is modelled.
This fact is what gives GPs their flexibility---they can model any time
series with a similar covariance structure.
In addition, a very simple function can usually capture the covariance
structure of a light curve, whereas a much more complicated one may be
required to model the time-series itself.
% The light curve of \Kepler-452 has been modelled with a GP in figure
% \ref{fig:GP_example}, shown in orange.
In figure \ref{fig:GP_example} the light curve of KIC 1430163 has been
modelled with a GP.
% Both provide adequate fits to the data, however only the periodic kernel
% function, `QP' is a useful model because it has a periodic parameter.
% I return to this point shortly.

A range of kernel functions could be used to describe stellar variability.
For example, the simplest and most commonly used kernel function, the `Squared
Exponential' (SE) produces an adequate fit to the KIC 1430163 light curve.
The SE kernel function is defined as,
\begin{equation}
k_{i,j} = A \exp \left(-\frac{(x_i - x_j)^2}{2l^2} \right).
\end{equation}
\label{eq:SE}
Here $A$ is the amplitude of covariance, $l$ is the length scale of
exponential decay and $x_i-x_j$ is the separation between data points.
The SE kernel function has the advantage of being very simple---it has just
two parameters, a covariance amplitude and length scale: $A$ and $l$.
If $l$ is large two data points far apart in $x$ will be tightly correlated,
and if small they will be loosely correlated.
Another property of the SE kernel function is that it produces functions that
are infinitely differentiable.
It is therefore possible to model a data set and its derivatives
simultaneously.
% \citep[see \eg][]{Rajpaul2015}
The SE kernel function may be an adequate model of the covariance in stellar
light curves, but it is not a {\it useful} one for the problem of rotation
period inference because it does not have a parameter that describes a period.
In order to infer rotation periods it is necessary to use a periodic kernel
function.
For this reason, we use the `Quasi-Periodic' kernel function.
The QP kernel function is defined as
\begin{equation}
k_{i,j} = A \exp \left(-\frac{(x_i - x_j)^2}{2l^2} -
\Gamma^2 \sin^2(\frac{2\pi}{P}) \right).
\end{equation}
\label{eq:QP}
It is the product of the SE kernel function, which describes the overall
covariance decay, and an exponentiated, squared, sinusoidal kernel function
that describes the periodic covariance structure.
$P$ can be interpreted as the rotation period of the star and $\Gamma$
controls the amplitude of the sinusoidal term.
This kernel function allows two data points that are separated in time by one
rotation period to be tightly correlated, while points separated by half a
period can be weakly correlated.
Both the QP kernel function was used to produce the
model shown in figure ~\ref{fig:GP_example}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/001430163.pdf}
\caption[A light curve with a GP model.]
{Light curve of KIC 1430163, an active star with a rotation period of $\sim$ 4
days.
The blue solid line shows a fit to the data using
a Gaussian process model with a QP covariance kernel function and the pink
broken line shows a fit with a SE kernel function.
Both kernel functions provide an adequate fit to the data, however we use the
QP function because it has a useful parameter that corresponds to the star's
rotation period.}
\label{fig:GP_example}
\end{center}
\end{figure}
